---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("tools", "README-"),
  dev = if (require("Cairo")) "CairoPNG" else "png"
)
```

# qualpalr

<!-- badges: start -->
[![R-CMD-check](https://github.com/jolars/qualpalr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jolars/qualpalr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/jolars/qualpalr/branch/master/graph/badge.svg)](https://app.codecov.io/gh/jolars/qualpalr?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/qualpalr)](https://cran.r-project.org/package=qualpalr)
<!-- badges: end -->

`qualpalr` generates distinct qualitative color palettes, primarily for use in
R graphics. Given `n` (the number of colors to generate), along with a subset in
the [hsl color space](https://en.wikipedia.org/wiki/HSL_and_HSV) (a cylindrical
representation of the RGB color space) `qualpalr` attempts to find the `n`
colors in the provided color subspace that *maximize the smallest pairwise color
difference*. This is done by projecting the color subset from the HSL color
space to the DIN99d space. DIN99d is (approximately) perceptually uniform, that
is, the euclidean distance between two colors in the space is proportional to
their perceived difference.

`qualpalr` was inspired by
[i want hue](http://medialab.github.io/iwanthue/).

## Usage

Let's create a palette of 4 colors of hues from 0 to 360, saturation between
0.1 and 0.5, and lightness from 0.6 to 0.85.

```{r pal1}
library(qualpalr)

pal <- qualpal(n = 4, list(h = c(0, 360), s = c(0.1, 0.5), l = c(0.6, 0.85)))
```

We can retrieve the generated colors in hex format

```{r hex}
pal$hex
```

Alternatively, we can create a palette using one of the predefined color
subspaces and, optionally, adapt it to color deficiency (here *deuteranomaly*).

```{r pal2}
pal2 <- qualpal(
  n = 4,
  colorspace = "pretty",
  cvd = "deutan",
  cvd_severity = 0.5
)
```

We can look at a distance matrix of the pairwise color differences from the 
palette we have generated.

```{r dedin99d}
pal2$de_DIN99d
```

The palette may also be plotted with in a multidimensional scaling plot
generated from the distance matrix of the colors in the palette.

```{r plot, fig.width=5, fig.height=5}
plot(pal2)
```

Or it might be plotted in one of the provided color spaces as a
scatterplot matrix.

```{r pairs, fig.width=5, fig.height=5}
pairs(pal2, colorspace = "HSL")
```

## Installation

The current CRAN release can be installed by running

```{r install_cran, eval = FALSE}
install.packages("qualpalr")
```

The development version can be installed by running

```{r install_github, eval = FALSE}
devtools::install_github("jolars/qualpalr")
```

## Versioning

Versioning is based on [semantic versioning](https://semver.org/). 

## Code of conduct

Please note that this project is released with a
[Contributor Code of Conduct](https://github.com/jolars/qualpalr/blob/master/CONDUCT.md). By participating in this project you
agree to abide by its terms.

## License

`qualpalr` is open source software, licensed under [GPL-3](https://github.com/jolars/qualpalr/blob/master/LICENSE).

