---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.width = 5,
  fig.height = 5
)
```

# qualpalr

[![Build Status](https://travis-ci.org/jolars/qualpalr.svg?branch=master)](https://travis-ci.org/jolars/qualpalr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jolars/qualpalr?branch=master&svg=true)](https://ci.appveyor.com/project/jolars/qualpalr)
[![Coverage Status](https://codecov.io/github/jolars/qualpalr/coverage.svg?branch=master)](https://codecov.io/github/jolars/qualpalr?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/qualpalr)](http://cran.r-project.org/package=qualpalr)

`qualpalr` generates distinct qualitative color palettes, primarily for use in R graphics. The user inputs `n` (the number of colors to generate), along with a subset in the [hsl color space](https://en.wikipedia.org/wiki/HSL_and_HSV) (a cylindrical representation of the RGB color space). Using a [Nelder-Mead optimizer](https://en.wikipedia.org/wiki/Nelder%E2%80%93Mead_method), `qualpalr` then attempts to find the `n` colors in the provided color subspace that *maximize the smallest pairwise color difference*, computed via the [CIEDE2000 color difference formula](https://en.wikipedia.org/wiki/Color_difference#CIEDE2000).

`qualpalr` was inspired by [i want hue](http://tools.medialab.sciences-po.fr/iwanthue/).

## Usage

```{r titanic}
library(qualpalr)

# Create a palette of 4 colors of hues from 0 to 360, saturations between
# 0.1 and 0.5, and lightness from 0.6 to 0.85
pal <- qualpal(n = 4, list(h = c(0, 360), s = c(0.1, 0.5), l = c(0.6, 0.85)))

# Look at the colors in hex format
pal$hex

# Create a palette using one of the predefined color subspaces
pal2 <- qualpal(n = 4, colorspace = "colorblind")

# Produce a distance matrix of the CIEDE2000 color differences
pal2$ciede2000
```

Methods for `pairs()` and `plot()` have been prepared for objects created with `qualpal`. 

```{r plot_methods, eval = FALSE}
plot(pal)
pairs(pal, colorspace = "hsl")
```

## Installation

The current CRAN release can be installed by running

```{r install_cran, eval = FALSE}
install.packages("qualpalr")
```

The development version can be installed by running

```{r install_github, eval = FALSE}
devtools::install_github("jolars/qualpalr")
```

## License

`qualpalr` is open source software, licensed under GPL-3.

