---
title: "Comparisons to Alternatives"
author: "Johan Larsson"
date: "`r Sys.Date()`"
bibliography: qualpal.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to qualpalr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

Trying to find a good color palette for categorical data is a problem
that has been tackled previously. In this vignette, we will compare
the qualpal library with other alternatives, namely

- colorgogical
- palletailor
- distinctipy
  glasbey
- I want Hue


```{r}
library(qualpalr)

hcl_space <- list(h = c(50, 290), c = c(0, 134), l = c(25, 85))

pal <- qualpal(n = 5, hcl_space)

colorgogical <- c("#b4ddd4", "#9b233d", "#2eece6", "#f82387", "#32a190")

iwanthue <- c(
  "#5f683c",
  "#5c87b5",
  "#79bf4b",
  "#c79041",
  "#7bb99d"
)

qualpal_result <- analyze_palette(pal$hex)$normal
analyze_palette(colorgogical)$normal
analyze_palette(iwanthue)$normal
```

## Palettailor

```{r}
palettailor <- c(
  "#4ca600",
  "#5dfbbf",
  "#eda59f",
  "#32998d",
  "#605ec4",
  "#0099ef",
  "#e950e8",
  "#e1f32e",
  "#936411",
  "#e9003e"
)

hsl_space <- list(h = c(0, 360), s = c(0, 1), l = c(0.45, 0.95))

pal <- qualpal(10, hsl_space)

analyze_palette(palettailor)$normal
analyze_palette(pal$hex)$normal
```

## Color Vision Deficiency 

Some of the tools above feature capabilities to adapt palettes to color
vision deficiencies. 
