---
title: "An introduction to qualpalr"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to qualpalr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`R` contains several tools that produce high-quality sequential and diverging color palettes, but few that are dedicated to qualitative color palettes. ´qualpalr´ is an attempt to fill that gap by providing an easy-to-use interface to generate qualitative color palettes of any size and from whatever range of colors the user prefers. 

Provided `n` (the desired number of colors to generate), and a subset of the [hsl](https://en.wikipedia.org/wiki/HSL_and_HSV) color space, `qualpalr` uses a [Nelder-Mead optimizer](https://en.wikipedia.org/wiki/Nelder%E2%80%93Mead_method) (`nmkb` from [dfoptim](https://cran.r-project.org/web/packages/dfoptim/index.html) to find the configuration of `n` colors that maximize the smallest pairwise [CIEDE2000](https://en.wikipedia.org/wiki/Color_difference#CIEDE2000) color difference.

## Examples

`qualpalr` relies on one basic function: `qualpal()` which takes as its input `n` (the number of colors to generate) and `colorspace`, which can be either

* a list of numeric vectors `h` (hue from -360 to 360), `s` (saturation from 0 to 1), and `l` (lightness from 0 to 1), all of length 2, specifying a min and max.
* a character vector specifying one of the predefined color subspaces, which at the time of writing are *pretty*, *pretty_dark*, *colorblind*, *rainbow*, and *pastels*.

```{r basic_usage}
library(qualpalr)
set.seed(1) # The optimizer uses random initial parameters
pal <- qualpal(n = 5, list(h = c(0, 360), s = c(0.4, 0.6), l = c(0.5, 0.85)))
pal <- qualpal(n = 4, colorspace = "pretty")
```

The resulting object, `pal2`, is a list with several color tables and a distance matrix based on the CIEDE2000 formula.

```{r qualpal_list}
pal
```

For more descriptive versions of what colors were generated, methods for `pairs` and `plot` have been written for `qualpal` objects:

```{r fig.show='hold'}
# Multidimensional scaling plot
plot(pal)

# Pairs plot in the Lab color space
pairs(pal, colorspace = "Lab")

```

The colors are most easily used in R by accessing `pal$hex`

```{r map, fig.width = 5, fig.height = 5}
library(maps)
map("france", fill = TRUE, col = pal$hex, mar = c(0, 0, 0, 0))
```

## Details

Control parameters can be passed to `nmkb` via `...` in `qualpal`, of which `maxfeval` (default: 5000) and `tol` (default: 1e-6) are probably of primary interest. See the documentation to [dfoptim](https://cran.r-project.org/web/packages/dfoptim/index.html) for details.

`qualpalr` was not designed for speed. It is slow, particularly for large `n`, and I *strongly* recommend aginst using it in other functions. 
